---
output: html_document
editor_options: 
  chunk_output_type: console
---
# Occupancy {#occupancy}

```{r setup08, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

Occupancy modelling has been one of the mainstays of camera traps data analysis for many years, so learning how to wangle our data into occupancy-style formats is essential. 

When we survey wild and free ranging populations using any sampling methodology, the probability of detecting a given individual or species if it is actually present on the landscape at the time of sampling is typically less than  one. This is because wild animals are often hard to see! This issue is termed "imperfect detection". 

In order to deal with the imperfect detection issue - occupancy models separate our the counts of a given species at a site into two processes: 

1) occupancy (ψ) - which is the probability of a species occurring within a spatial unit (or “site”) during the sampling session
2) detection probability (p) - the probability that the species will be detected given that it already occurs at a site 

In order to separate out the occupancy process from the detection process, surveys need to occur at replicated 'sites' and we need repeated 'visits' to the same site. It is important to know that in camera trap studies, practitioners typically treat individual locations as sites and rather than repeated return to a location to survey it at different times, they divide the continuous camera activity data into block of time (e.g. 1 to 7 day windows).   

Occupancy models were not developed specifically for camera traps - thus there are a suite of assumptions we need to make about the populations we survey when applying occupancy models. We do not adress these here. However, below we provide a list introductory resources for you to dig into the occupancy models to decide if they are appropriate for your situation:

[Burton, A. Cole, et al. "Wildlife camera trapping: a review and recommendations for linking surveys to ecological processes." Journal of Applied Ecology 52.3 (2015): 675-685.](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2664.12432)

[MacKenzie, Darryl I., et al. Occupancy estimation and modeling: inferring patterns and dynamics of species occurrence. Elsevier, 2017.](https://pubs.er.usgs.gov/publication/5200296)

Let's focus our time on getting our data into the right formt, and applying some occupancy models!


```{r ch8_1, echo=T, results='hide', message =F, warning=F , class.source="Rmain"}
# Check you have them and load them
list.of.packages <- c("kableExtra", "tidyr", "ggplot2", "gridExtra", "dplyr", "unmarked", "lubridate", "tibble", "sf", "gfcanalysis", "MuMIn", "spOccupancy")

new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
lapply(list.of.packages, require, character.only = TRUE)

```


## Single species occupancy model

In this example we will use the `...weekly_observations` dataframe we created in the [data creation](#data-creation) section. We do this because 7 days is a time interval which occupancy models are often devided into for occupancy analyses.  

We first need to create a site by occasion matrix for our focal species, using a 7-day occasion length. This means we need to break our camera data into seven day bins. 

```{r ch8_2, class.source="Rmain"}
# Import the weekly observations data set
week_obs <- read.csv("data/processed_data/AlgarRestorationProject_30min_independent_weekly_observations.csv", header=T)

# Use white-tailed deer as an example
focal_sp<- "Odocoileus.virginianus"

# subset to just the 2019 data
tmp_week <- week_obs[substr(week_obs$date,1,4)==2019,]

# Create the Y data  
y_dat <- tmp_week[,c("placename", "date", focal_sp)] %>% # Subset to just white-tailed deer
            pivot_wider(names_from = date, values_from = focal_sp) # Shift to wide format

# Convert it to a matrix - but only keep the date values
y_mat <- as.matrix(y_dat[,unique(tmp_week$date)])

# Update the row names
row.names(y_mat) <- y_dat$placename
```

The resulting data frame looks like this:

```{r  ch8_6, echo=F}
kbl(y_mat)%>%
  kable_paper() %>%
  scroll_box( height = "200px")
```

It is a matrix of all the weeks the cameras were active, and whether the count of the independent detections in that interval. The `fill = NA` command puts a zero where there is data for a given day. 

You can see that in some columns we have values > 1 - this is because we had more than one independent observation in that week. Occupancy analyses (typically) require this data to be in detection/non-dection (0 or 1) format. So lets change that here and add back in our effort data.

```{r ch8_7, class.source="Rmain"}
# Where y_mat is > 1, and where y_mat isn't NA - give it the value 1
y_mat[y_mat>1 & is.na(y_mat)==F] <- 1

# Effort matrix
# To create the effort matrix - inst of the Focal Species bring in the effort
eff_mat <- tmp_week[,c("placename", "date", "days")]

eff_mat <-  eff_mat %>%
  # Create a matrix based on dates and effort
  spread(date,days, fill = NA) %>% 
  # group by deloyment Location ID, then make that the row.namesd
  group_by(placename) %>%
  column_to_rownames( var = "placename") 

eff_mat <- as.matrix(eff_mat)

# Remove all of the data from the weeks where we did not get a complete sample:
y_mat[eff_mat!=7] <- NA

```

Now we are ready to feed this into the `unmarked` package. 

### Unmarked package

One of the hurdles in using the `unmarked` package is it uses a different style of dataframe called an unmarked dataframe. It is essentially a compillation of the different dataframes we need for the analysis (y data and covariate data). We asemmbled the Y data above, so now lets make the covariates:

```{r ch8_11, class.source="Rmain"}
locs <-  read.csv("data/processed_data/AlgarRestorationProject_camera_locations_and_covariates.csv")

# Unmarked wants your detection history, effort data and site covariates as matrices. But the order is important!
# Check the order of your matrices and covariates files matches... or you will get nonsense!
table(locs$placename == row.names(y_mat))

# Standardise you explanatory variables
library(MuMIn)
z_locs <- stdize(locs)

# Build an unmarkedFramOccu
un_dat <- unmarkedFrameOccu(y = y_mat, # your occupancy data
                            siteCovs = z_locs) # Your site covariates 


# Fit a basic model 
m0 <- occu(formula = ~1 # detection formula first
                     ~1, # occupancy formula second,
                data = un_dat)

# View the results
summary(m0)
```

The estimate you see for both occupancy and detection probability is on the log-link scale. If we want to calculate the occupancy probability, we can use the `backTransform()` function:

```{r ch8_16, class.source="Rmain"}
backTransform(m0, type = "state")
```

So the probability that a white-tailed deer occupies one of the survey locations is ~0.66. For the detection probability we specify "det":

```{r ch8_17, class.source="Rmain"}
backTransform(m0, type = "det")

```

The probability that we detect a white-tailed deer in a given unit of time (7-days), given that it is there to be detected, is ~0.2.

Let's fit a couple of other models!

First with a continuous covariate on the occupancy probability, then a categorical one too:

```{r ch8_18, class.source="Rmain"}
# Occupancy is influence by line of sight
m1 <- occu(formula = ~1 # detection formula first
                     ~z.line_of_sight_m, # occupancy formula second,
                data = un_dat)

# Occupancy is influenced by the feature_type a camera is deployed on
m2 <- occu(formula = ~1 # detection formula first
                     ~feature_type, # occupancy formula second,
                data = un_dat)

# Perform model selection to find the best one
model.sel(m0,m1,m2)
```

The best supported model contains `z.line_of_sight_m`, although the improvement on the null model is minimal. 

### Plotting predictions

We can observe the relationship between our covariates and our occupancy probabilities through the use of a dummy dataframes (which we will call `new_dat`). A dummy dataframe is essential just a dataframe built up of `dummy` data - which lies within the upper and lower limits of the covariates we already have. We wouldn't want to extrapolate beyond our data! We can then plot the results:

```{r ch8_20, class.source="Rmain"}
# Generate new data to predict from 
new_dat <- cbind(expand.grid(
                  z.line_of_sight_m=seq(min(z_locs$z.line_of_sight_m),max(z_locs$z.line_of_sight_m), # add more covariates here if the model is more complex
                  length.out=25)))

# Make the predicted values for the data you supplied                 
new_dat <- predict(m1, type="state", newdata = new_dat, appendData=TRUE)


#Plot the results

p1 <- ggplot(new_dat, aes(x = z.line_of_sight_m, y = Predicted)) + # mean line
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.5, linetype = "dashed") + #Confidence intervals
  geom_path(linewidth = 1) +
  labs(x = "Line of sight", y = "Occupancy probability") + # axis labels
  theme_classic() +
  coord_cartesian(ylim = c(0,1))

p1
```

As with our habitat use model, white-tailed deer (**Odocoileus virginianus**) occupancy appears to decrease with increasing line of sight.  

### On your own

Let's explore some of the models we fit in the [habitat use chapter](#habitat-use) in the occupancy framework. We have not included any detection covariates in this example dataset, so hold that constant for now!


